\begin{thebibliography}{9}
\addcontentsline{toc}{part}{Bibliography}

\bibitem{Christie2015}
  Thony Christie, %Author
  \emph{Calendrical confusion or just when did Newton die?}, %Title
  % Edition (if applicable)
  \url{https://thonyc.wordpress.com/2015/03/20/calendrical-confusion-or-just-when-did-newton-die/}, % URL (if applicable)
  The Renaissance Mathematics, % Journal/Website Title (if applicable)
  % Volume (if applicable)
  % Number (if applicable)
  % Pages (if applicable)
  % Publisher
  2015. % Year

\bibitem{Eklund1996}
  Matti Eklund, %Author
  \emph{On How Logic Became First Order}, %Title
       % Edition (if applicable)
  Nordic Journal of Philosophical Logic, % Journal (if applicable)
  Vol. 1, % Volume (if applicable)
  No. 2, % Number (if applicable)
  pp. 147-167, % Pages (if applicable)
  Scandinavian University Press, % Publisher
  1996. % Year
  
  \bibitem{Enderton1977}
  H. B. Enderton, %Author
  \emph{Elements of Set Theory}, %Title
       % Edition (if applicable)
  % URL (if applicable)
  % Journal/Website Title (if applicable)
  % Volume (if applicable)
  % Number (if applicable)
  % Pages (if applicable)
  Academic Press, % Publisher
  1977. % Year
  
  \bibitem{Haaparanta2009}
  Leila Haaparanta, %Author
  \emph{The Development of Modern Logic}, %Title
       % Edition (if applicable)
  % URL (if applicable)
  % Journal/Website Title (if applicable)
  % Volume (if applicable)
  % Number (if applicable)
  % Pages (if applicable)
  Oxford University Press, % Publisher
  2009. % Year
  
  \bibitem{HuthRyan2004}
  M. Huth, and M. Ryan, %Author
  \emph{Logic in Computer Science: Modelling and Reasoning about Systems}, %Title
       % Edition (if applicable)
  % URL (if applicable)
  % Journal/Website Title (if applicable)
  % Volume (if applicable)
  % Number (if applicable)
  % Pages (if applicable)
  Cambridge University Press, % Publisher
  2004. % Year

\bibitem{Kleene1952}
  Stephen C. Kleene, %Author
  \emph{Introduction to Metamathematics}, %Title
   % Edition
  North-Holland Publishing Co, % Publisher
  1952. % Year

\bibitem{Mendelson2010}
  Elliott Mendelson, %Author
  \emph{Introduction to Mathematical Logic}, %Title
  5th edition, % Edition
  CRC Press, % Publisher
  2010. % Year
  
  \bibitem{MerzbackBoyer2011}
  U.C. Merzbach and C. B. Boyer, %Author
  \emph{A History of Mathematics}, %Title
  3rd edition, % Edition
  Wiley, % Publisher
  2011. % Year
  
  \bibitem{RMRobinson1950}
  R. M. Robinson, %Author
  \emph{An Essentially Undecidable Axiom System}, %Title
       % Edition (if applicable)
  % URL (if applicable)
  Proceedings of the International Congress of Mathematicians, % Journal/Website Title (if applicable)
  % Volume (if applicable)
  % Number (if applicable)
  pp. 729 - 730, % Pages (if applicable)
  % Publisher
  1950. % Year

\end{thebibliography}